---
title: "Comparison distance-based versus model-based approach"
author: "Aurélien Boyé"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmdformats::readthedown:
    use_bookdown: TRUE
    number_sections : FALSE
    fig_caption: TRUE
    code_folding: "hide"
    highlight: "tango"
    lightbox: TRUE
    gallery: TRUE
    thumbnails: FALSE # if TRUE display content images as thumbnails
    toc_depth: 6
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE, cache = TRUE, fig.cap = "\\s")
```

# Packages {.tabset}

## Data handling

```{r}
library(tidyverse)

# Plot
library(ggrepel)
library(patchwork)
library(ggboral) # devtools::install_github("mbedward/ggboral")
```


## Analysis

```{r}

# Distance-based
#---------------

library(vegan)
#library(adespatial)

# JSDM
#------

library(mvabund)
library(boral)

```

## Custom functions

```{r}

#--------------------------------------------------
# Plot ordinations from rda(), cca(), or cmdscale()
#--------------------------------------------------

plot_ord <- function(ord,comm, reverse_axis2 = FALSE){
  
  site_scores <- scores(ord, display = "sites") %>%
    as.data.frame() %>%
    set_names("Axis1", "Axis2")
  
  if(reverse_axis2){
    site_scores$Axis2 <- -site_scores$Axis2 
  }
  
  if(any(class(ord) %in% c("rda","cca"))){
    
   sp_scores <- scores(ord, display = "sp") %>%
     as.data.frame() %>%
    set_names("Axis1", "Axis2")
    
  }else{# PCoA
    
    sp_scores <- wascores(site_scores, comm) %>%
      as.data.frame()
  }
 
  ggplot() +
    geom_point(data = site_scores, aes(x = Axis1, y = Axis2)) +
    geom_path(data = site_scores, aes(x = Axis1, y = Axis2)) +
    geom_text_repel(data = site_scores, aes(x = Axis1, y = Axis2, label = rownames(site_scores)), nudge_y = -0.1) +
    geom_segment(data = sp_scores, aes(x = 0, y = 0, xend = Axis1, yend = Axis2), col = "grey") +
    geom_text_repel(data = sp_scores, aes(x = Axis1, y = Axis2, label = rownames(sp_scores)), col = "grey") +
    coord_fixed(ratio = 1) +
    theme_minimal()
   
}


#--------------------------------------------------
# Plot ordinations from boral
#--------------------------------------------------

plot_boral <- function(boral_mod){
  
  # Retrieve data to do the same plots as before
  ord_data <- gg_lvsplot(boral_mod) %>%
    layer_data(., 2) %>%
    select(x, y, label) %>%
    mutate(group = if_else(str_detect(label, "sp"), "species", "sites"))
  
   ggplot() +
    geom_point(data = ord_data %>% filter(group == "sites"), aes(x = x, y = y)) +
    geom_path(data = ord_data %>% filter(group == "sites"), aes(x = x, y = y)) +
    geom_text_repel(data = ord_data %>% filter(group == "sites"), aes(x = x, y = y, label = label), nudge_y = -0.1) +
    geom_segment(data = ord_data %>% filter(group == "species"), aes(x = 0, y = 0, xend = x, yend = y), col = "grey") +
    geom_text_repel(data = ord_data %>% filter(group == "species"), aes(x = x, y = y, label = label), col = "grey") +
    coord_fixed(ratio = 1) +
    theme_minimal() +
     xlab("Latent variable 1") +
     ylab("Latent variable 2")
   
}

```


---

# Legendre et Gallagher 2001

## Create data of Fig. 3a


```{r}

fig3a_data <- data.frame(
  site = seq(1:19),
  sp1 = c(7,4,2,1, rep(0,15)),
  sp2 = c(1,2,4,7,8,7,4,2,1, rep(0,10)),
  sp3 = c(rep(0,5),1,2,4,7,8,7,4,2,1,rep(0,5)),
  sp4 = c(rep(0,10),1,2,4,7,8,7,4,2,1),
  sp5 = c(rep(0,15), 1,2,4,7),
  sp6 = c(0,1,1,rep(0,16)),
  sp7 = c(rep(0,6),2,1,rep(0,11)),
  sp8 = c(rep(0,11),3,1,rep(0,6)),
  sp9 = c(rep(0,16), 4,1,0)
)

pos <- position_dodge(0.3)

fig3a_data %>%
  gather(species, abundance, -site) %>%
  ggplot(data= ., aes(x = site, y = abundance, fill = species, shape = species, group = species)) +
  geom_line(aes(col = species), position = pos, alpha = 0.6) +
  geom_point(data = . %>% filter(abundance != 0), alpha=0.6, size = 2.5, position = pos) +
  scale_shape_manual(values=rep(c(21:25),4)) +
  scale_x_continuous(breaks = seq(1:19)) +
  theme_linedraw()

fig3a_data <- fig3a_data %>%
  column_to_rownames("site")
```

## Reproduce Fig. 4


> Using decostand :
>
> - profile transformation: Y.tr = decostand(Y, "total")
> - chord transformation: Y.tr = decostand(Y, "norm")
> - log-chord transformation: Y.tr = decostand(log1p(Y), "norm")
> - Hellinger transformation: Y.tr = decostand(Y, "hellinger")
> - chi-square transformation: Y.tr = decostand(Y, "chi.sq")
>
> > [dist.ldc](https://rdrr.io/cran/adespatial/man/dist.ldc.html) can also be used

```{r}

comm <- fig3a_data

# Figure 4a : PCA on raw data
#-----------------------------

ord <- rda(comm)
fig_4a <- plot_ord(ord, comm) +
  ggtitle("PCA on raw data")

# Figure 4b : CA on raw data
#---------------------------

ord <- cca(comm)
fig_4b <- plot_ord(ord, comm) +
  ggtitle("CA on raw data")

# Figure 4c : PCA on chi2-transformed data
#-----------------------------------------

comm_tr <- decostand(comm,"chi.sq")
ord <- rda(comm_tr)
fig_4c <- plot_ord(ord, comm) +
  ggtitle("PCA on chi2-transformed data")

# Figure 4d : PCA on chord-transformed data
#------------------------------------------

comm_tr <- decostand(comm,"norm")
ord <- rda(comm_tr)
fig_4d <- plot_ord(ord, comm) +
  ggtitle("PCA on chord-transformed data")

# Figure 4e : PCA on species-profiles transformed data
#-----------------------------------------------------

comm_tr <- decostand(comm,"total")
ord <- rda(comm_tr)
fig_4e <- plot_ord(ord, comm) +
  ggtitle("PCA on species-profiles transformed data")

# Figure 4f : PCA on Hellinger transformed data
#---------------------------------------------

comm_tr <- decostand(comm,"hellinger")
ord <- rda(comm_tr)
fig_4f <- plot_ord(ord, comm) +
  ggtitle("PCA on Hellinger-transformed data")

# Figure 4g : PCoA with bray-curtis distance
#---------------------------------------------

comm_d <- vegdist(comm, "bray")
ord <- cmdscale(comm_d)
fig_4g <- plot_ord(ord, comm, reverse_axis2 = TRUE) +
  ggtitle("PCoA with Bray-curtis")

```

```{r, fig.width = 15, fig.height = 15}
(fig_4a + fig_4b + fig_4c) / 
  (fig_4d + fig_4e + fig_4f) / 
  (fig_4g + plot_spacer() + plot_spacer()) +
  plot_annotation(tag_levels = 'a', tag_prefix="(", tag_suffix = ")")
```

> - Add variance explained to axes labels
> - Do log-chord

## Comparison of Fig. 4 with JSDM

### Comparison with boral

#### Normal

```{r, results = 'hide'}
# Normal - without row effects
#----------------------------

boral_normal <- boral(y = comm, family = "normal", lv.control = list(num.lv = 2), row.eff = "none")

boral_p1 <- plot_boral(boral_normal)+
  scale_y_reverse() +
  scale_x_reverse() +
  ggtitle("Normal - without row effects")

# Normal - fixed row effects
#----------------------------

boral_normal_fixed <- boral(y = comm, family = "normal", lv.control = list(num.lv = 2), row.eff = "fixed")

boral_p2 <- plot_boral(boral_normal_fixed)+
  scale_y_reverse() +
  ggtitle("Normal - fixed row effects")

# Normal - random row effects
#----------------------------

boral_normal_rand <- boral(y = comm, family = "normal", lv.control = list(num.lv = 2), row.eff = "random")

boral_p3 <- plot_boral(boral_normal_rand)+
  scale_y_reverse() +
  scale_x_reverse() +
  ggtitle("Normal - random row effects")

# Hellinger - Normal - without row effects
#----------------------------
comm_tr <- decostand(comm,"hellinger")

boral_normal_hell <- boral(y = comm_tr, family = "normal", lv.control = list(num.lv = 2), row.eff = "none")

boral_p4 <- plot_boral(boral_normal_hell)+
  scale_y_reverse() +
  scale_x_reverse() +
  ggtitle("Hellinger-transformed - Normal - without row effects")

# Normal - fixed row effects
#----------------------------

boral_normal_hell_fixed <- boral(y = comm_tr, family = "normal", lv.control = list(num.lv = 2), row.eff = "fixed")

boral_p5 <- plot_boral(boral_normal_hell_fixed)+
  scale_y_reverse() +
  ggtitle("Hellinger-transformed - Normal - fixed row effects")

# normal - random row effects
#----------------------------

boral_normal_hell_rand <- boral(y = comm_tr, family = "normal", lv.control = list(num.lv = 2), row.eff = "random")

boral_p6 <- plot_boral(boral_normal_hell_rand)+
  scale_y_reverse() +
  scale_x_reverse() +
  ggtitle("Hellinger-transformed - Normal - random row effects")

```

```{r, fig.width = 8, fig.height = 12}
(boral_p1 + boral_p4) / (boral_p2 + boral_p5) / (boral_p3 + boral_p6) +
  plot_annotation(tag_levels = 'a', tag_prefix="(", tag_suffix = ")")
```

```{r}
(boral_p1 + boral_p2 + boral_p3)+
  plot_annotation(tag_levels = 'a', tag_prefix="(", tag_suffix = ")")
```


```{r, eval = FALSE, include = FALSE, fig.width = 15, fig.height = 8}
(boral_p1 + boral_p2 + boral_p3) / (boral_p4 + boral_p5 + boral_p6) +
  plot_annotation(tag_levels = 'a', tag_prefix="(", tag_suffix = ")")
```

#### Poisson & Negative binomial

```{r, results = 'hide'}

# Poisson - without row effects
#----------------------------

boral_poisson <- boral(y = comm, family = "poisson", lv.control = list(num.lv = 2), row.eff = "none")

boral_p1 <- plot_boral(boral_poisson)+
  scale_y_reverse() +
  scale_x_reverse() +
  ggtitle("Poisson - without row effects")

# Poisson - fixed row effects
#----------------------------

boral_poisson_fixed <- boral(y = comm, family = "poisson", lv.control = list(num.lv = 2), row.eff = "fixed")

boral_p2 <- plot_boral(boral_poisson_fixed)+
  scale_y_reverse() +
  ggtitle("Poisson - fixed row effects")

# Poisson - random row effects
#----------------------------

boral_poisson_rand <- boral(y = comm, family = "poisson", lv.control = list(num.lv = 2), row.eff = "random")

boral_p3 <- plot_boral(boral_poisson_rand)+
  scale_y_reverse() +
  scale_x_reverse() +
  ggtitle("Poisson - random row effects")

# Negative binomial - without row effects
#--------------------------------------

boral_negbin <- boral(y = comm, family = "negative.binomial", lv.control = list(num.lv = 2), row.eff = "none")

boral_p4 <- plot_boral(boral_negbin) +
  scale_y_reverse() +
  ggtitle("Negative binomial - without row effects")

# Negative binomial - fixed row effects
#--------------------------------------

boral_negbin_fixed <- boral(y = comm, family = "negative.binomial", lv.control = list(num.lv = 2), row.eff = "fixed")

boral_p5 <- plot_boral(boral_negbin_fixed) +
  scale_y_reverse() +
  ggtitle("Negative binomial - fixed row effects")

# Negative binomial - random row effects
#--------------------------------------

boral_negbin_rand<- boral(y = comm, family = "negative.binomial", lv.control = list(num.lv = 2), row.eff = "random")

boral_p6 <- plot_boral(boral_negbin_rand) +
  scale_y_reverse() +
  ggtitle("Negative binomial - random row effects")

```

```{r, fig.width = 8, fig.height = 12}
(boral_p1 + boral_p4) / (boral_p2 + boral_p5) / (boral_p3 + boral_p6) +
  plot_annotation(tag_levels = 'a', tag_prefix="(", tag_suffix = ")")
```

#### Residuals of the models

- **Normal - without row effects**

```{r}
plot(boral_normal, ask = F, mfrow = c(2,2))
```

- **Poisson - without row effects**

```{r}
plot(boral_poisson, ask = F, mfrow = c(2,2))
```

- **Negative binomial - without row effects**

```{r}
plot(boral_negbin, ask = F, mfrow = c(2,2))
```

### Comparison with mvabund

```{r}

meanvar.plot(comm)

```

## Comparison with Ecocopula ?

https://cran.r-project.org/web/packages/ecoCopula/vignettes/the_basics.html

---

# Session info

```{r session_info}

sessionInfo()

```